Regression Cube Analysis of Cohort Study Data

### 2015-03-31 draft.tex
cleaned up comments

%% Abstract section.
\abstract{%
%\com{Problem}.
Epidemiological studies comprise heterogeneous data about a subject group (a \emph{cohort}) to define disease-specific risk factors.
%%
These data contain information (\emph{features}) about a subject's lifestyle, medical condition as well as medical image data.
%%
%These features are analyzed using statistical regression analysis to identify feature combinations indicating a disease (the \emph{target feature}).
%%
Statistical regression analysis is used to evaluate these features and to identify feature combinations indicating a disease (the \emph{target feature}).
%%
Although there is a strong demand for overview visualizations of a whole data set towards a target feature, no suitable tool is available for epidemiological researchers.

@@

Testing features for association with diseases using regression analysis is one of the most important epidemiological tools.
%%
%Using it to assess the statistical resilience of a hypothesis rarely involves \emph{more than three features} due to the required subject count.
Using it to assess the statistical resilience of a hypothesis rarely involves \emph{more than three features} due to the higher dimensional problem and the required subject count.
%%
% missing overview techniques ist zu hart; es gibt ja techniken (eine scatterplot matrix zb.)
Due to the amount of data and only limited overview techniques, possible correlations may be missed.
%%
Explorative analyses and overview visualizations of the data set as presented in prior work \cite{Klemm2014VIS} are not tailored to a specific target feature and mostly highlight correlations between features, which are known to the domain expert (e.g., correlation between body size and spine shape).
%%
We incorporate the regression analysis, which is familiar to the domain experts, into overview visualizations to support a hypothesis-free analysis or an analysis towards a specific disease or hypothesis.
%%
This is achieved by providing template regression formulas, which are applied to all potential feature combinations.
%%
Since the notation is familiar to epidemiologists, they can rapidly include their domain knowledge into the analysis process.
%%
Difference views between regression formulas allow to assess the influences of individual features on the process.

Our contributions are:
\begin{itemize}
	\item An overview visualization technique representing feature interactions using target features.
	\item Visualization techniques, which incorporate overview visualizations of all regression analyses at once as well as details on demand techniques for in-depth investigations of feature relationships.
	\item Freely adjustable regression formulas provide a simple, yet powerful way to adjust the regression analysis to specific hypotheses about the data.
	\item The analysis of confounding features by providing comparison views between different formula results.
%	\item The open and web-based approach of the system allows for analysis of any data using the presented method.
	\item The expansion and adaption of our method to any data with our open web-based system.
\end{itemize}
%%
The term `Regression Cube' has been used for the detailed analysis of one or a small number of regression models \cite{Ahmadi, Chan}.
%%
They extend 2D scatter plots to display detailed information about a regression model.
%%
In behalf of that, we declare our visualization technique for comparing and analyzing a large number of different regression models with different input features as \emph{Regression Cube}.
%%
\section{Epidemiological Background} \label{sec:Background}
%%
This section covers the epidemiological workflow and requirements.
%%
\subsection{Epidemiological Workflow} \label{EpidemiologicalWorkflow}
Epidemiological research is performed by experts from different academic disciplines, such as physicians, statisticians and medical computer scientists focusing on biometrics and image segmentation.
%%
Their goal is to derive disease-specific risk factors by assessing epidemiological features with statistical methods.
%%
As described by Thew et al.~\cite{Thew2009} the epidemiological workflow is divided into these different steps:
\begin{itemize}
	\item Clinicians make observations in the daily practice, which are translated into hypotheses.
	\item Epidemiologists compile a list of features depicting the hypothesis and include confounding features.
	\item Statisticians assess the association of the derived features towards the investigated disease.
\end{itemize}
%%
Relative risks can be determined if a statistical resilient association of features towards a condition is extracted.
%%
They indicate the per-subject chance of developing the disease.
%%
Reproducibility of results is an epidemiological key requirement and guides all analyses steps.
%%
Statistical programs, such as \texttt{SPSS}, are used to analyze the data regarding the classical sequential epidemiological workflow. Hence, images are mostly used for the communication of results, rather than for providing insight into the data.
%%

In \cite{Klemm2014VIS} we described an Interactive Visual Analysis approach for image-centric cohort study data, which connects to the feature listing step.
%%
The methods aim to derive hypotheses through analysis of the data and observations about previously unknown feature correlations.
%%
Employing \emph{hypothesis generation} requires overview visualizations of feature correlations, which are not supported by standard statistical processors.
%%
In this work, we focus on a similar approach to derive insight and even new hypotheses through the data rather than only using it for a confirmatory analysis.
%An alternative data-driven approach is described as follows.
%%
% \paragraph{Data Driven Hypothesis-Generation-based Analysis.}
% We already described an Interactive Visual Analysis approach for image-centric cohort study data, which connects to the variable listing step \cite{Klemm2014VIS}.
% %%
% The methods aim to derive hypotheses through analysis of the data and observations about previously unknown feature correlations.
% %%
% Employing \emph{hypotheses generation} requires overview visualizations of feature correlations, which are not supported by standard statistical processors.

\subsection{Epidemiological Data} \label{sec:EpidemiologicalData}
%%
%\com{Bernhard: Due to overlaps to VIS'14 paper in this section, this can be shortened. E.g. "More discussion of data types and characteristics can be found in [...]"}
%%
There are many data acquisition modalities available to epidemiologists, which are suitable for different analysis types.
%%
%We focus on cohort studies are a major epidemiological tool for gathering epidemiological features.
We focus on cohort studies, which impose the largest data sets and yield a highly heterogenous and incomplete information space.
%%
Subject data are collected either towards a specific disease or with the widest range possible.
%%
The latter allows the data set to be assessed towards different diseases.
%%
The feature space comprises information about lifestyle, somatometric features, medical parameters, genetic data as well as medical images.
%%
These features are derived through different modalities, such as questionnaires, medical examinations or laboratory analyses.
%%
Many features are sparse, such as follow-up questions about a medication or treatment of a certain disease.
%%
Other features are exclusive for a sub-group, such as women-specific questions, e.g., number of born children or period status.
%%
%\com{Restriction of amount of data because it must not be triangulated and because the ethics committees are very restrictive}
%%
%\paragraph{Data Types.}
%%
Medical status features or lifestyle factors are often of \emph{dichotomous} (binary) type.
%%
%The data space comprises continuous variables (somatometric variables, such as body weight) as well as categorical variables (e.g. graduation type).
%%
%The data heterogeneity has to be taken into account when the analysis method is chosen.
%%
Continuous data are often discretized (e.g. 10 year steps for age) to equalize the feature types and to simplify the method selection.
%%
However, this should be avoided, since it reduces the information space and also introduces a new information bias, as assumptions are modeled through the discretization.
%%

Medical image data is also acquired and analyzed in modern cohort studies.
%%
We incorporate image-derived features, but do not focus on analyzing image data.
%%
More discussions of cohort study data types and characteristics can be found in \cite{Preim2014, Toennies2015}.
%%
%\paragraph{Image Data.}
% \com{Can probably be cut (reference to Klaus' Paper for more details on this) (see Bernhards comment above)}
% %%
% Since the Rotterdam study, many modern cohort studies include medical image data.
% %%
% For ethical reasons, the imaging modalities usually not include ionizing radiation.
% %%
% The image quality is often inferior to clinical standards, which is a tradeoff between time and cost \cite{Preim2014}.
% %%
% These data are hard to analyze as they require segmentations to outline structures of interest.
% %%
% This process is prone to inter- and intra-observer variability when carried out manually.
% %%
% Automatic or semi-automatic solutions are a remedy for this problem, but are costly and need to be tailored to the structure of interest \cite{Toennies2015}.
% %%
% Image-derived variables are either dichotomous variables indicating a medical finding by a radiologist or continuous, describing the shape of segmented structures (e.g. volume or diameter).
\\\\
%\paragraph{Confounding Features.}
Features influencing the exposure as well as the outcome of an analysis are called \emph{confounders} and have to be specially treated.
%Confounding features have to be considered in epidemiological analysis, as they influence both the exposure as well as the outcome of an analysis.
%%
%Features influencing the exposure as well as the outcome of an analysis are called confounders.
%%
The analysis model has to be adjusted by normalizing all included features towards the confounder.
%%
\emph{Age} is included as confounder in almost any epidemiological analysis, since most diseases (such as different cancer types) are more likely with increasing age.
%%
It also influences the general body condition and thereby almost all features acquired through cohort studies.
%%
Another important confounder is \emph{gender}.
%%
Other confounders have to be selected by epidemiologists specific to the investigated condition.

@@

\paragraph{Regression Analysis Notation.} Regression formulas are usually denoted as follows:
%%
\begin{equation}
Dependent \sim Independent_1 + ... + Independent_n
%Dependent \sim \sum_{i=1}^n Independent_i
\label{eq:RegressionNotation}
\end{equation}
%%
An example of a regression formula would be $KidneyDisorder \sim Smoking + Obesity$.
The most commonly used regression operators comprise:
%%
\begin{itemize}
	\item $+,-$ inclusion/exclusion of the variable (e.g. $x\pm y$),
	\item $:$ inclusion of interactions between the variables (e.g. $x:y$),
	\item $*$ inclusion of the variables as well as their interactions (e.g. $x*y$)
	\item $|$ (conditioning) inclusion of variable x, given y (e.g. $x|y$)
\end{itemize}
%%
The class of the target feature restricts the regression type.
%%
Different regression types are available, we focus on the following:
%%
\paragraph{Linear Regression for Continuous Target.} The basic type is the linear regression, creating a linear map from the space comprising the \emph{independent} features to the \emph{dependent} features.
%%
The \emph{dependent} variable has to be of a continuous type.
%%

\paragraph{Logistic Regression for Dichotomous Target.} Logistic regression implies a dichotomous target variable.
%%
The target is described by fitting a logistic function.
%%
Logistic models, as opposed to linear models, do not allow for extracting an $R^2$ quality of fit value.
%%
Therefore, pseudo-$R^2$ values are extracted, such as the \emph{Nagelkerke $R^2$}, which mimics the behavior of the $R^2$.
%%
\emph{Nagelkerke $R^2$} behaves different than $R^2$ values extracted from the linear regression model.
%%
Comparisons have to be handled with care.

%\com{Regression cubes are an important tool for checking statistical resilience.}
%\com{Describe how regression cubes work (including notation).}
%\com{Move this to Regression Cube part?}

\subsection{The Study of Health in Pomerania (SHIP)}
The \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%%
It does not focus on a specific disease, making the data set open for many diseases.
%%
Unique for the \texttt{SHIP} is the acquisition of medical image data per cohort subject.
%%
A second cohort, \texttt{SHIP-TREND} was started in 2012.
%%
Data for both cohorts are examined in a 5-year time span.
%%
New parameters are added in each iteration, extending the range of investigated diseases.
%%
For the latest two acquisitions (\texttt{SHIP-2} and \texttt{SHIP-TREND-0}), MRI scans are included into the cohort \cite{Hegenscheid2009, Ivanovska2014}.
%%
However, most examinations occur in all stages and are performed according to the same instructions to enable comparisons over the different stages of the study. % ("\emph{method loyality}").
%%
%Starting in 1997 with a cohort of 4,308 subjects, the \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%After the pioneering Rotterdam study (started in 1990), several MR imaging study initiatives were initiated.
%
%They slightly differ in clinical focus, acquired data and epidemiological research questions.
%
%Starting in 1997 with a cohort of 4,308 subjects, the \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%
%Data are collected without focus on a group of diseases.
%
%This allows to query the data regarding many diseases and conditions.
%
%
\section{Prior and Related Work}
%% https://en.wikipedia.org/wiki/Exploratory_data_analysis#cite_note-Tukey1977-4
In 1977, Tukey already stated that data is too often only analyzed using confirmatory data analysis \cite{Tukey}.
%%
He emphasized the need to use data to derive hypotheses instead, which can then be tested again.
%%
In this section we present prior and related work trying to achieve this goal.
%%
%We put each paper in relation to our work, derive ideas and point out differences, starting with the works closest to ours:
%%
\\\\
The work of Piringer et al.~\cite{Piringer}, Chan et al.~\cite{Chan}, Angelelli et al.~\cite{Angelelli} and Zhang et al.~\cite{Zhang2014} are closest related to ours regarding different aspects.
%% HyperMoVal: Interactive Visual Validation of Regression Models for Real-Time Simulation
Akin to our approach, Piringer et al.~\cite{Piringer} propose methods for visualizing regression analysis results and properties for developing car engines.
%%
Their main goal is to assess the pairwise influence of independent features towards the target feature using a plot matrix displaying models as contours.
%%
They also incorporate 3D visualizations for each pairwise combination, but mainly because of its popularity with the target domain engineers.
%%
Linked views of model deviations allow to select outliers.
%%
This limits the method to comparing a few models at once as the plot matrix gets complex with increasing feature number.
%%
The main difference is their focus on analyzing one complex model in detail, yielding extensive plots, while we analyze a large amount of simpler models in terms of \emph{different} features.
%%
They also focus on metric features, while we process categorical data as well.
%%
%We share the matrix-arrangement of results depicting relationships between independent features in a model.

%% Regression Cube: A Technique for Multidimensional Visual Exploration and Interactive Pattern Finding
We use the term `Regression Cube' for a comparative 3D representation of multiple regression models.
%% einheitlich 2D-XXX oder 2D XXX, meist ist es 2D XXX
Chan et al.~\cite{Chan} use the same term to describe an extension of the 2D scatter plot representation of a linear regression model (incorporating solely metric features) to a 3D Cube.
%%
They group subjects using a set of interaction techniques as well as clustering algorithms to calculate sub-groups, which can then be compared using their cube representation.
%%
Similar to \cite{Piringer}, they focus on highlighting details of the included models rather than comparing models consisting of different features.
%%
Insight is derived by subject grouping, which spawns new cube correlations and therefore allows drilling down to the data.
%%
In contrast to this approach, we focus on comparing models using quality-of-fit measures and do not focus on subdividing the data sets to gain insight.
%%
We model expert knowledge through the definition of regression formulas, which is not the focus of Chan et al.
%%

@@

\paragraph{Visual Analysis in Public Health.}
%%--
Shneiderman et al.~\cite{Shneiderman2013} highlight the necessity of interactive visualizations in healthcare data. %, e.g. using \emph{Health 2.0}, which is an umbrella for web-based health-informatics frameworks.
%%
%They divide Health 2.0 in three domains, \emph{Personal health information}, \emph{clinical health information}, and \emph{public health information}.
%%
%We focus on the latter.
%%
%Shneiderman and colleagues also present challenges,
The presented challenges in this application comprise the systems to be sufficiently comprehensible and intuitive to fit into the short time span that clinicians have for analyzing data.
%%
Furthermore, the systems should provide comparative relationship visualization, characterization of similarity and presenting risks.
%%
%We use the challenges as design guidelines for our web-based prototype, focusing on functional clean designs to allow fast insight into relationships towards diseases.
%%
%%

%%
In their survey on interactive information visualizations for health record data, Rind et al.~\cite{Rind} identified future work addressing the usability of such frameworks.
%%
%We follow their recommendation in designing an open system available to everyone, applicable to a various data sets.
%%
We follow their recommendation to design an open system available to everyone and applicable to various data sets.
%%
Schreck and Keim \cite{Schreck} present a method for analyzing social media data with focus on characterization and cause of an epidemic.
%%
They employ data aggregation techniques and use geographical data to map social media events, focusing on word clouds to derive insight into possible causes.
%%
They derive hypotheses using their high variation average message density algorithms, steering the user's focus to particular events.
%%
%The focus remains on geolocated social media data rather than subjects with multiple dimensions.
%%
%These and similar systems follow the same analysis work as we do to derive insight into data.

@@

While the presented techniques were well received by the epidemiological experts, the explanatory power towards back pain was limited, yielding an analysis based on image-derived features, such as extracting and analyzing curvature, torsion and angle of the lumbar spine \cite{Klemm2015}.
%%
%These measures were analyzed using a novel decision tree quality plot towards non-image data.
%%
We concluded that the model quality is insufficient to characterize back pain.
\\\\
The difference of the presented approach towards our previous and related work is twofold:
\begin{enumerate}
	\item We focus on the large-scale analysis of a vast number of linear and logistic regression-models by assessing their quality-of-fit using descriptive metrics.
	\item The analysis is conducted w.r.t. a target feature and incorporates expert knowledge via the regression model definition rather than subdividing the underlying data.
\end{enumerate}
%%
\section{Regression Cube Analysis of Cohort Study Data}
%%
The basic idea of our \emph{Regression Cube} is to provide an overview visualization of large cohort study data sets towards target features.
%%
Overview visualizations of feature relationships as presented by Angelelli et al.~\cite{Angelelli} are often focused on relationships between the visualized features.
%%
Correlation metrics, such as the \emph{Pearson product-moment correlation coefficient} or \emph{Cram\'{e}r's V} contingency values are incorporated to achieve this goal.
%%
In epidemiology, these relationships are also of interest, but rather w.r.t. their explanatory power towards the target feature.
%%
These target features often indicate the presence of the investigated disease.
%%
As described in Section \ref{sec:RegressionAnalysis}, regression analysis is the statistical tool of choice for analyzing these relationships.
%%
% unified -> singular [Steffi?]
A regression model is based on expert knowledge; there is no singular rule how to apply models to a given set of features. Thus, they have to be applied with care.
%%
\subsection{Cube Description Using Regression Formula Notation}
%%
Expert knowledge is integrated into regression analysis by the regression formulas.
%%
As described in Section~\ref{sec:RegressionAnalysis}, the formula input influences the type of the chosen regression method as well as the \emph{independent} features describing the target.
%%
%In order to calculate all combinations of two to three independent features, we can apply  \emph{Regression Cube} aims to
%%

Since we want to associate the regression analyses with an overview visualization, we are interested in all possible combinations of (two or more) independent features describing a target.
%%
We achieve this by introducing dynamic variables $X$, $Y$ and $Z$ into the regression notation.
%%
Our method replaces the dynamic variables with all features in the data set.
%%
In a data set with $n$ (e.g., 100) features, the regression formula
\begin{equation}
Cancer \sim X + Y
\end{equation}
would yield $n^2$ (10,000) regression models, describing all possible combinations of two features in the data describing the target feature $Cancer$.
%%
The major advantage of this notation is that it comes natural to anyone familiar with regression analysis, since it is the standard way of expression.
%%
%This allows for a fast adaptation and in the epidemiological application domain.
%%
With simple adjustments to the formula, different results can be achieved:
%%
\begin{itemize}
	\item $Z \sim X + Y$ calculates all combinations of two features w.r.t. all possible target features.
	\item $Cancer \sim X + Y + BodyWeight$ includes the $BodyWeight$ feature into all regression models as feature with $Cancer$ as target.
	\item $Cancer \sim X + Y + Z$ calculates all combinations of three features towards the $Cancer$ target.
\end{itemize}
%%
The problem with this approach lies in its complexity.
%%
The number of calculated regression models exponentially increases for each dynamic variable added.
%%
If we assume a data set with 100 features and the formula $Z \sim X + Y$, we obtain 1,000,000 regression models.
%%
% hours -> h da ms ja auch benutzt wird.
When each regression takes about 50~ms calculation, we roughly have to wait 14~h for the calculation to complete.
%%
Therefore, the computational complexity needs to be reduced.
%%
An approach for this is presented in the following section.
%\com{Regression formulas are described using standard regression formula notation by introducing variables.}
%%
%\com{Target Group: Physicians/Epidemiologists with basic statistical background to understand regression formulas and Interaction.}
%%
\subsection{Target-Variable-Dependent Dimension Reduction}
% \com{Uli, Can you please proofread this?}
% %%
% The vast majority of features in an epidemiological data have no or a very low relation towards the target feature.
% %%
% Identifying these features und excluding them from the calculation can reduce the number of dimension significantly.
% %%

>>>>

%% Abstract section.
\abstract{%
%\com{Problem}.
Epidemiological studies comprise heterogeneous data about a subject group (a \emph{cohort}) to define disease-specific risk factors.
%%
These data contain information (\emph{features}) about a subject's lifestyle, medical condition as well as medical image data.
%%
Statistical regression analysis is used to evaluate these features and to identify feature combinations indicating a disease (the \emph{target feature}).
%%
Although there is a strong demand for overview visualizations of a whole data set towards a target feature, no suitable tool is available for epidemiological researchers.

@@

Testing features for association with diseases using regression analysis is one of the most important epidemiological tools.
%%
Using it to assess the statistical resilience of a hypothesis rarely involves \emph{more than three features} due to the higher dimensional problem and the required subject count.
%%
Due to the amount of data and only limited overview techniques, possible correlations may be missed.
%%
Explorative analyses and overview visualizations of the data set as presented in prior work \cite{Klemm2014VIS} are not tailored to a specific target feature and mostly highlight correlations between features, which are known to the domain expert (e.g., correlation between body size and spine shape).
%%
We incorporate the regression analysis, which is familiar to the domain experts, into overview visualizations to support a hypothesis-free analysis or an analysis towards a specific disease or hypothesis.
%%
This is achieved by providing template regression formulas, which are applied to all potential feature combinations.
%%
Since the notation is familiar to epidemiologists, they can rapidly include their domain knowledge into the analysis process.
%%
Difference views between regression formulas allow to assess the influences of individual features on the process.

Our contributions are:
\begin{itemize}
	%%
	\item An overview visualization technique representing feature interactions using target features.
	%%
	\item Visualization techniques, which incorporate overview visualizations of all regression analyses at once as well as details on demand techniques for in-depth investigations of feature relationships.
	%%
	\item Freely adjustable regression formulas provide a simple, yet powerful way to adjust the regression analysis to specific hypotheses about the data.
	%%
	\item The analysis of confounding features by providing comparison views between different formula results.
	%%
	\item The expansion and adaption of our method to any data with our open web-based system.
	%%
\end{itemize}
%%
The term `Regression Cube' has been used for the detailed analysis of one or a small number of regression models \cite{Ahmadi, Chan}.
%%
They extend 2D scatter plots to display detailed information about a regression model.
%%
In behalf of that, we declare our visualization technique for comparing and analyzing a large number of different regression models with different input features as \emph{Regression Cube}.
%%
\section{Epidemiological Background} \label{sec:Background}
%%
This section covers the epidemiological workflow and requirements.
%%
\subsection{Epidemiological Workflow} \label{EpidemiologicalWorkflow}
Epidemiological research is performed by experts from different academic disciplines, such as physicians, statisticians and medical computer scientists focusing on biometrics and image segmentation.
%%
Their goal is to derive disease-specific risk factors by assessing epidemiological features with statistical methods.
%%
As described by Thew et al.~\cite{Thew2009} the epidemiological workflow is divided into these different steps:
\begin{itemize}
	\item Clinicians make observations in the daily practice, which are translated into hypotheses.
	\item Epidemiologists compile a list of features depicting the hypothesis and include confounding features.
	\item Statisticians assess the association of the derived features towards the investigated disease.
\end{itemize}
%%
Relative risks can be determined if a statistical resilient association of features towards a condition is extracted.
%%
They indicate the per-subject chance of developing the disease.
%%
Reproducibility of results is an epidemiological key requirement and guides all analyses steps.
%%
Statistical programs, such as \texttt{SPSS}, are used to analyze the data regarding the classical sequential epidemiological workflow. Hence, images are mostly used for the communication of results, rather than for providing insight into the data.
%%

In \cite{Klemm2014VIS} we described an Interactive Visual Analysis approach for image-centric cohort study data, which connects to the feature listing step.
%%
The methods aim to derive hypotheses through analysis of the data and observations about previously unknown feature correlations.
%%
Employing \emph{hypothesis generation} requires overview visualizations of feature correlations, which are not supported by standard statistical processors.
%%
In this work, we focus on a similar approach to derive insight and even new hypotheses through the data rather than only using it for a confirmatory analysis.
%%
\subsection{Epidemiological Data} \label{sec:EpidemiologicalData}
%%
There are many data acquisition modalities available to epidemiologists, which are suitable for different analysis types.
%%
We focus on cohort studies, which impose the largest data sets and yield a highly heterogenous and incomplete information space.
%%
Subject data are collected either towards a specific disease or with the widest range possible.
%%
The latter allows the data set to be assessed towards different diseases.
%%
The feature space comprises information about lifestyle, somatometric features, medical parameters, genetic data as well as medical images.
%%
These features are derived through different modalities, such as questionnaires, medical examinations or laboratory analyses.
%%
Many features are sparse, such as follow-up questions about a medication or treatment of a certain disease.
%%
Other features are exclusive for a sub-group, such as women-specific questions, e.g., number of born children or period status.
%%
Medical status features or lifestyle factors are often of \emph{dichotomous} (binary) type.
%%
Continuous data are often discretized (e.g. 10 year steps for age) to equalize the feature types and to simplify the method selection.
%%
However, this should be avoided, since it reduces the information space and also introduces a new information bias, as assumptions are modeled through the discretization.
%%

Medical image data is also acquired and analyzed in modern cohort studies.
%%
We incorporate image-derived features, but do not focus on analyzing image data.
%%
More discussions of cohort study data types and characteristics can be found in \cite{Preim2014, Toennies2015}.
%%
\\\\
Features influencing the exposure as well as the outcome of an analysis are called \emph{confounders} and have to be specially treated.
%%
The analysis model has to be adjusted by normalizing all included features towards the confounder.
%%
\emph{Age} is included as confounder in almost any epidemiological analysis, since most diseases (such as different cancer types) are more likely with increasing age.
%%
It also influences the general body condition and thereby almost all features acquired through cohort studies.
%%
Another important confounder is \emph{gender}.
%%
Other confounders have to be selected by epidemiologists specific to the investigated condition.

@@

\paragraph{Regression Analysis Notation.} Regression formulas are usually denoted as follows:
%%
\begin{equation}
Dependent \sim Independent_1 + ... + Independent_n
\label{eq:RegressionNotation}
\end{equation}
%%
An example of a regression formula would be $KidneyDisorder \sim Smoking + Obesity$.
The most commonly used regression operators comprise:
%%
\begin{itemize}
	\item $+,-$ inclusion/exclusion of the variable (e.g. $x\pm y$),
	\item $:$ inclusion of interactions between the variables (e.g. $x:y$),
	\item $*$ inclusion of the variables as well as their interactions (e.g. $x*y$)
	\item $|$ (conditioning) inclusion of variable x, given y (e.g. $x|y$)
\end{itemize}
%%
The class of the target feature restricts the regression type.
%%
Different regression types are available, we focus on the following:
%%
\paragraph{Linear Regression for Continuous Target.} The basic type is the linear regression, creating a linear map from the space comprising the \emph{independent} features to the \emph{dependent} features.
%%
The \emph{dependent} variable has to be of a continuous type.
%%

\paragraph{Logistic Regression for Dichotomous Target.} Logistic regression implies a dichotomous target variable.
%%
The target is described by fitting a logistic function.
%%
Logistic models, as opposed to linear models, do not allow for extracting an $R^2$ quality of fit value.
%%
Therefore, pseudo-$R^2$ values are extracted, such as the \emph{Nagelkerke $R^2$}, which mimics the behavior of the $R^2$.
%%
\emph{Nagelkerke $R^2$} behaves different than $R^2$ values extracted from the linear regression model.
%%
Comparisons have to be handled with care.
%%
\subsection{The Study of Health in Pomerania (SHIP)}
The \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%%
It does not focus on a specific disease, making the data set open for many diseases.
%%
Unique for the \texttt{SHIP} is the acquisition of medical image data per cohort subject.
%%
A second cohort, \texttt{SHIP-TREND} was started in 2012.
%%
Data for both cohorts are examined in a 5-year time span.
%%
New parameters are added in each iteration, extending the range of investigated diseases.
%%
For the latest two acquisitions (\texttt{SHIP-2} and \texttt{SHIP-TREND-0}), MRI scans are included into the cohort \cite{Hegenscheid2009, Ivanovska2014}.
%%
However, most examinations occur in all stages and are performed according to the same instructions to enable comparisons over the different stages of the study.
%%
\section{Prior and Related Work}
%% https://en.wikipedia.org/wiki/Exploratory_data_analysis#cite_note-Tukey1977-4
In 1977, Tukey already stated that data is too often only analyzed using confirmatory data analysis \cite{Tukey}.
%%
He emphasized the need to use data to derive hypotheses instead, which can then be tested again.
%%
In this section we present prior and related work trying to achieve this goal.
%%
\\\\
The work of Piringer et al.~\cite{Piringer}, Chan et al.~\cite{Chan}, Angelelli et al.~\cite{Angelelli} and Zhang et al.~\cite{Zhang2014} are closest related to ours regarding different aspects.
%% HyperMoVal: Interactive Visual Validation of Regression Models for Real-Time Simulation
Akin to our approach, Piringer et al.~\cite{Piringer} propose methods for visualizing regression analysis results and properties for developing car engines.
%%
Their main goal is to assess the pairwise influence of independent features towards the target feature using a plot matrix displaying models as contours.
%%
They also incorporate 3D visualizations for each pairwise combination, but mainly because of its popularity with the target domain engineers.
%%
Linked views of model deviations allow to select outliers.
%%
This limits the method to comparing a few models at once as the plot matrix gets complex with increasing feature number.
%%
The main difference is their focus on analyzing one complex model in detail, yielding extensive plots, while we analyze a large amount of simpler models in terms of \emph{different} features.
%%
They also focus on metric features, while we process categorical data as well.
%%

%% Regression Cube: A Technique for Multidimensional Visual Exploration and Interactive Pattern Finding
We use the term `Regression Cube' for a comparative 3D representation of multiple regression models.
%%
Chan et al.~\cite{Chan} use the same term to describe an extension of the 2D scatter plot representation of a linear regression model (incorporating solely metric features) to a 3D Cube.
%%
They group subjects using a set of interaction techniques as well as clustering algorithms to calculate sub-groups, which can then be compared using their cube representation.
%%
Similar to \cite{Piringer}, they focus on highlighting details of the included models rather than comparing models consisting of different features.
%%
Insight is derived by subject grouping, which spawns new cube correlations and therefore allows drilling down to the data.
%%
In contrast to this approach, we focus on comparing models using quality-of-fit measures and do not focus on subdividing the data sets to gain insight.
%%
We model expert knowledge through the definition of regression formulas, which is not the focus of Chan et al.
%%

@@

\paragraph{Visual Analysis in Public Health.}
%%--
Shneiderman et al.~\cite{Shneiderman2013} highlight the necessity of interactive visualizations in healthcare data. %, e.g. using \emph{Health 2.0}, which is an umbrella for web-based health-informatics frameworks.
%%
%They divide Health 2.0 in three domains, \emph{Personal health information}, \emph{clinical health information}, and \emph{public health information}.
%%
%We focus on the latter.
%%
%Shneiderman and colleagues also present challenges,
The presented challenges in this application comprise the systems to be sufficiently comprehensible and intuitive to fit into the short time span that clinicians have for analyzing data.
%%
Furthermore, the systems should provide comparative relationship visualization, characterization of similarity and presenting risks.
%%
%We use the challenges as design guidelines for our web-based prototype, focusing on functional clean designs to allow fast insight into relationships towards diseases.
%%

In their survey on interactive information visualizations for health record data, Rind et al.~\cite{Rind} identified future work addressing the usability of such frameworks.
%%
%We follow their recommendation in designing an open system available to everyone, applicable to a various data sets.
%%
We follow their recommendation to design an open system available to everyone and applicable to various data sets.
%%
Schreck and Keim \cite{Schreck} present a method for analyzing social media data with focus on characterization and cause of an epidemic.
%%
They employ data aggregation techniques and use geographical data to map social media events, focusing on word clouds to derive insight into possible causes.
%%
They derive hypotheses using their high variation average message density algorithms, steering the user's focus to particular events.
%%
%The focus remains on geolocated social media data rather than subjects with multiple dimensions.
%%
%These and similar systems follow the same analysis work as we do to derive insight into data.

@@

While the presented techniques were well received by the epidemiological experts, the explanatory power towards back pain was limited, yielding an analysis based on image-derived features, such as extracting and analyzing curvature, torsion and angle of the lumbar spine \cite{Klemm2015}.
%%
%These measures were analyzed using a novel decision tree quality plot towards non-image data.
%%
We concluded that the model quality is insufficient to characterize back pain.
\\\\
The difference of the presented approach towards our previous and related work is twofold:
%%
\begin{enumerate}
	%%
	\item We focus on the large-scale analysis of a vast number of linear and logistic regression-models by assessing their quality-of-fit using descriptive metrics.
	%%
	\item The analysis is conducted w.r.t. a target feature and incorporates expert knowledge via the regression model definition rather than subdividing the underlying data.
	%%
\end{enumerate}
%%
\section{Regression Cube Analysis of Cohort Study Data}
%%
The basic idea of our \emph{Regression Cube} is to provide an overview visualization of large cohort study data sets towards target features.
%%
Overview visualizations of feature relationships as presented by Angelelli et al.~\cite{Angelelli} are often focused on relationships between the visualized features.
%%
Correlation metrics, such as the \emph{Pearson product-moment correlation coefficient} or \emph{Cram\'{e}r's V} contingency values are incorporated to achieve this goal.
%%
In epidemiology, these relationships are also of interest, but rather w.r.t. their explanatory power towards the target feature.
%%
These target features often indicate the presence of the investigated disease.
%%
As described in Section \ref{sec:RegressionAnalysis}, regression analysis is the statistical tool of choice for analyzing these relationships.
%%
A regression model is based on expert knowledge; there is no singular rule how to apply models to a given set of features. Thus, they have to be applied with care.
%%
\subsection{Cube Description Using Regression Formula Notation}
%%
Expert knowledge is integrated into regression analysis by the regression formulas.
%%
As described in Section~\ref{sec:RegressionAnalysis}, the formula input influences the type of the chosen regression method as well as the \emph{independent} features describing the target.
%%
